---
title: "MutationalPatterns-SNVs"
---

```{r setup, include=FALSE}
library(dplyr)
library(MutationalPatterns)
library(BSgenome) 
library(gridExtra)
library(cowplot)
library(ggplot2)
library(GenomicRanges)
library(ggsci)
library(wesanderson)
library(colorspace)
library(ComplexHeatmap)
library(patchwork)

setwd("")
dir=""
plot_dir=""
vcf_dir="/path/to/vcfs/concatenated/per/mouse"

ref_genome <- "BSgenome.Mmusculus.UCSC.mm10"
library(ref_genome, character.only = TRUE)
```


```{r}
mouse_vcf_list<-list.files(vcf_dir, pattern = "vcf.gz", full.names = TRUE)
sample_names <- gsub("_concat\\.vcf\\.gz$", "", basename(mouse_vcf_list))

# Keep a named index for tracking
sample_index <- setNames(seq_along(sample_names), sample_names)
print(sample_index)

##create the vcf object
mouse_vcf_obj <- read_vcfs_as_granges(mouse_vcf_list, sample_names, ref_genome, predefined_dbs_mbs = TRUE)

#Add group metadata
group<- c(rep("active", 3), rep("dormant",3), rep("old", 3), rep("young", 3)) 

mouse_vcf_obj@metadata$group<-as.factor(group)

saveRDS(mouse_vcf_obj, paste0(dir, "mouse_vcf_obj_YOLRC.rds"))

#Clean
rm(mouse_vcf_names, sample_names, i, mouse_vcf_list)
```

Mutation characteristics
```{r}
##Retrieve base substitution types from the VCF GRangesList object as “REF>ALT” 
mouse_muts<-mutations_from_vcf(mouse_vcf_obj[[1]])

#Extract the conventional substitutions (purine>anything)
mouse_types_sub<-mut_type(mouse_vcf_obj[[1]])

#Retrieve the context (trinucleotides) for all the samples of the vcf objects
mouse_context<-type_context(mouse_vcf_obj[[1]], ref_genome)
lapply(mouse_context, head, 12)

#Count the mutation type occurrences for all the vcf objects in the GRangesLists
mouse_mut_occurrences<-mut_type_occurrences(mouse_vcf_obj, ref_genome)

#Create the trinucl matrix
mouse_mut_matrix<-mut_matrix(vcf_list=mouse_vcf_obj, ref_genome = ref_genome)
saveRDS(mouse_mut_matrix, paste0(paste0(dir, "mouse_mut_matrix.rds")))

#Pool the groups for the trinucleotide plots
pooled_mouse_mut_matrix<-pool_mut_mat(mouse_mut_matrix, grouping = mouse_vcf_obj@metadata$group)
write.csv(pooled_mouse_mut_matrix, paste0(plot_dir, "mouse_mut_matrix_pergroup.csv"))

YvsO_trinucl<-plot_96_profile(pooled_mouse_mut_matrix[, c(3,4)], condensed = TRUE)

CycvsDorm_trinucl<-plot_96_profile(pooled_mouse_mut_matrix[, c(1,2)], condensed = TRUE)
```

Strict refitting -- According to the authors of the package, the refit can be accurate with as little as 50 mutations per signature.
```{r}
#Impport COSMIC signatures
known_signatures<-get_known_signatures()
#Add the HSPC signature
stored_sigs <- read.table(paste0(dir, "ref_SBS_signatures_incl_hspc.txt"), header = T)
sigs_hspc <- cbind(stored_sigs$HSPC, known_signatures)
colnames(sigs_hspc)[1] <- 'HSPC'
known_signatures<-sigs_hspc

mouse_strict_refit <- fit_to_signatures_strict(mouse_mut_matrix, known_signatures, max_delta = 0.35)
mouse_strict_fit_res<- mouse_strict_refit$fit_res
plot_contribution(mouse_strict_fit_res$contribution, coord_flip = FALSE, mode = "relative")
```

Try refitting with the "best subset" method as well as it can be more accurate if we apply prior biological knowledge. 
eg. from the signatures that were chosen in the strict refitting above, I can filter out the ones that according to biology have no meaning in the dataset.
Try refitting with bootstrap, only for a subset of signatures that make sense
```{r}
selected_signatures<-known_signatures[,c(1, 2, 4, 6, 7, 12, 26, 39)] #Start with the signatures: HSPC, SBS1, SBS3, SBS5, SBS6, SBS8, SBS18, SBS32

mouse_bootstrap <- fit_to_signatures_bootstrapped(mouse_mut_matrix, selected_signatures, n_boots=150, method="strict")

plot_bootstrapped_contribution(mouse_bootstrap) #SBS3, SBS6, SBS32 are rarely called. I remove and re-run the bootstrapping. Remaining signatures are the ones usually reported in hematological datasets: HSPC, SBS1, SBS5, SBS18

selected_signatures<-known_signatures[,c(1, 2, 6, 26)]
mouse_bootstrap <- fit_to_signatures_bootstrapped(mouse_mut_matrix, selected_signatures, n_boots=150, method="strict")

plot_bootstrapped_contribution(mouse_bootstrap, mode = "relative") 
saveRDS(mouse_bootstrap, paste0(plot_dir, "mouse_bootstrap_object_150runs_strict.rds"))


plot_bootstrapped_contribution(mouse_bootstrap, mode = "relative", plot_type = "dotplot")


#Subset the mouse_bootstrap for the plot
young_old_indices = grep("young|old", rownames(mouse_bootstrap))

# Subset the matrix using the indices
YvsO_bootstrap = mouse_bootstrap[young_old_indices, ]
saveRDS(YvsO_bootstrap, paste0(plot_dir, "YvsO_mouse_bootstrap.rds"))

plot_bootstrapped_contribution(YvsO_bootstrap, mode="relative", plot_type = "dotplot")


ActivevsDormant_indices=grep("active|dormant", rownames(mouse_bootstrap))
ActivevsDormant_bootstrap = mouse_bootstrap[ActivevsDormant_indices, ]
saveRDS(ActivevsDormant_bootstrap, paste0(plot_dir, "ActivevsDormant_mouse_bootstrap_matrix.rds"))

plot_bootstrapped_contribution(CyclingvsDormant_bootstrap, mode="relative", plot_type = "dotplot")
```

Calculate the average relative contribution of each signature from the bootstraps
```{r}
row_total_strict<-rowSums(mouse_bootstrap)
mouse_bootstrap_relative_contributions <- sweep(mouse_bootstrap, 1, row_total_strict, FUN="/")

saveRDS(mouse_bootstrap_relative_contributions, paste0(plot_dir, "mouse_bootstrap_strict_matrix_relative_contributions.rds" ))

mouse_bootstrap_relative_contributions<-as.data.frame(mouse_bootstrap_relative_contributions)
mouse_bootstrap_relative_contributions$Sample<-rownames(mouse_bootstrap_relative_contributions)
mouse_bootstrap_relative_contributions$BaseSampleName <- gsub("_[^_]+$", "", mouse_bootstrap_relative_contributions$Sample)

mouse_bootstrap_average_contributions_strict <- mouse_bootstrap_relative_contributions %>%
  group_by(BaseSampleName) %>%
  summarise(across(-Sample, mean, na.rm = TRUE))

saveRDS(mouse_bootstrap_average_contributions_strict, paste0(plot_dir, "sigs_average_relative_contributions_strict.rds"))
```

Check the similarity between mutational profiles and signatures
```{r}
cos_sim(mouse_mut_matrix[,1], known_signatures[,1])

cos_sim_samples_signatures<-cos_sim_matrix(mouse_mut_matrix, selected_signatures)
cos_sim_samples_signatures[7:12, ]
saveRDS(cos_sim_samples_signatures, paste0(dir,"cosine_similarity_between_samples_selected_sigs_YOLRC.rds"))


plot_cosine_heatmap(cos_sim_samples_signatures, cluster_rows = FALSE, cluster_cols = FALSE)
```

Estimate the absolute SNV burden attributed to each signature
```{r}
contrib<-read_excel(paste0(plot_dir, "signature_contributions_allsamples.xlsx"))
contrib_yo<- contrib %>% filter(group %in% c("young", "old"))

SBS1_palette=c("#7FC6A4", "#D6F8D6")

pdf(file=paste0(plot_dir, "SBS1_mutations_boxplot_YvsO.pdf"), width = 2.5, height = 4)

sbs1<-ggplot(contrib_yo, aes(x = group, y = SBS1_mutations, fill=group)) + 
  geom_violin(trim=FALSE) +
  geom_boxplot(width=0.05, color="white")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1, alpha=0.7, color="grey18", fill="grey18")+
  theme_classic()+
  scale_fill_manual(values = SBS1_palette)+
  guides(fill="none", color="none")+
  geom_signif(comparisons = list(c("young", "old")), 
                   map_signif_level = TRUE)+
  ylab("SBS1 substitutions")

sbs1

dev.off()

wilcox_test_SBS1_muts_YvsO <- wilcox.test(SBS1_mutations ~ group, data = contrib_yo)
print(wilcox_test_SBS1_muts_YvsO)


HSPC_palette=c("#713E5A", "#CA6680")

pdf(file=paste0(plot_dir, "HSPC_mutations_boxplot_YvsO.pdf"), width = 2.5, height = 4)

hspc<-ggplot(contrib_yo, aes(x = group, y = HSPC_mutations, fill=group)) + 
  geom_violin(trim=FALSE) +
  geom_boxplot(width=0.05, color="white")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1, alpha=0.7, color="grey18", fill="grey18")+
  theme_classic()+
  scale_fill_manual(values = HSPC_palette)+
  guides(fill="none", color="none")+
  geom_signif(comparisons = list(c("young", "old")), 
                   map_signif_level = TRUE)+
  ylab("HSPC substitutions")

hspc
dev.off()

wilcox_test_hspc_muts_YvsO <- wilcox.test(HSPC_mutations ~ group, data = contrib_yo)
print(wilcox_test_hspc_muts_YvsO)


SBS5_palette=c("#E57A44", "#EDC79B")

pdf(file=paste0(plot_dir, "SBS5_mutations_boxplot_YvsO-20032024.pdf"), width = 2.5, height = 4)

sbs5<-ggplot(contrib_yo, aes(x = group, y = SBS5_mutations, fill=group)) + 
  geom_violin(trim=FALSE) +
  geom_boxplot(width=0.05, color="white")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1, alpha=0.7, color="grey18", fill="grey18")+
  theme_classic()+
  scale_fill_manual(values = SBS5_palette)+
  guides(fill="none", color="none")+
  geom_signif(comparisons = list(c("young", "old")), 
                   map_signif_level = TRUE)+
  ylab("SBS5 substitutions")

sbs5
dev.off()

wilcox_test_sbs5_muts_YvsO <- wilcox.test(SBS5_mutations ~ group, data = contrib_yo)
print(wilcox_test_sbs5_muts_YvsO)

SBS18_palette=c( "#247BA0", "#BCEDF6")

pdf(file=paste0(plot_dir, "SBS18_mutations_boxplot_YvsO-20032024.pdf"), width = 2.5, height = 4)
ggplot(contrib_yo, aes(x = group, y = SBS18_mutations, fill=group)) + 
  geom_violin(trim=FALSE) +
  geom_boxplot(width=0.05, color="white")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1, alpha=0.7, color="grey18", fill="grey18")+
  theme_classic()+
  scale_fill_manual(values = SBS18_palette)+
  guides(fill="none", color="none")+
  geom_signif(comparisons = list(c("young", "old")), 
                   map_signif_level = TRUE)+
  ylab("SBS18 substitutions")
dev.off()

wilcox_sbs18_muts_YvsO <- wilcox.test(SBS18_mutations ~ group, data = contrib_yo)
print(wilcox_test_sbs18_muts_YvsO)

#patchwork
hspc + sbs1 + sbs5 + sbs18

```

Linear regression to estimate the rate of acquisition of each signature
```{r}
#Add the "age" column 
contrib_yo <- contrib_yo %>% mutate(age = case_when(
   group == "young" ~ "8",  
    group == "old" ~ "24" 
  ))

#HSPC
model_hspc <- lm(HSPC_mutations ~ age, data = contrib_yo)
summary(model_hspc)

pdf(file=paste0(plot_dir, "HSPC_substitutions_LinRegr.pdf"), width = 4, height = 3)

hspc_LR<-ggplot(contrib_yo, aes(x = age, y = HSPC_mutations, color = age)) +
  geom_point(size=5, alpha=0.7) +                                  # Add points
  geom_smooth(method = "lm", se = TRUE, color="grey18", size=0.5)+
  theme_classic()+
  scale_color_gradient(low = "#713E5A", high = "#CA6680")+
  ylab("HSPC substitutions")+
  xlab("Age (in months)")+
  scale_x_continuous(breaks = seq(0, 24, by = 2), limits = c(0, 24))+
  ylim(0,100)+
  guides(color=FALSE)

hspc_LR

dev.off()


#SBS5
model_sbs5 <- lm(SBS5_mutations ~ age, data = contrib_yo)
summary(model_sbs5)

pdf(file=paste0(plot_dir, "SBS5_substitutions_LinRegr.pdf"), width = 4, height = 3)

sbs5_LR<-ggplot(contrib_yo, aes(x = age, y = SBS5_mutations, color = age)) +
  geom_point(size=5, alpha=0.7) +                                 
  geom_smooth(method = "lm", se = TRUE, color="grey18", size=0.5)+
  theme_classic()+
  scale_color_gradient(low = "#E57A44", high = "#EDC79B")+
  ylab("SBS5 substitutions")+
  xlab("Age (in months)")+
  scale_x_continuous(breaks = seq(0, 24, by = 2), limits = c(0, 24))+
  ylim(0,120)+
  guides(color=FALSE)

sbs5_LR

dev.off()
```

